---
title: "Machine Learning Question 1"
author: "Simge Cinar"
date: "2023-12-08"
output:
  pdf_document: 
    latex_engine: xelatex
    fig_width: 5.5
    fig_height: 3.5
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(geosphere)
library(knitr)
```

# Question 1: KERNEL METHODS
**Question:** Implement a kernel method to predict the hourly temperatures for a date and place in Sweden. To do so, you are provided with the files stations.csv and temps50k.csv. These files contain information about weather stations and temperature measurements in the stations at different days and times. The data have been kindly provided by the Swedish Meteorological and Hydrological Institute (SMHI). \
You are asked to provide a temperature forecast for a date and place in Sweden. The forecast should consist of the predicted temperatures from 4 am to 24 pm in an interval of 2 hours. Use a kernel that is the sum of three Gaussian kernels:

\begin{itemize}
  \item The first to account for the physical distance from a station to the point of interest. For this purpose, use the function distHaversine from the R package geosphere.
  \item The second to account for the distance between the day a temperature measurement was made and the day of interest.
  \item The third to account for the distance between the hour of the day a temperature measurement was made and the hour of interest.
\end{itemize}

Choose an appropriate smoothing coefficient or width for each of the three kernels above. No cross-validation should be used. Instead, choose manually a width that gives large kernel values to closer points and small values to distant points. Show this with a plot of the kernel value as a function of distance. Help: Note that the file temps50k.csv may contain temper- ature measurements that are posterior to the day and hour of your forecast. You must filter such measurements out, i.e. they cannot be used to compute the forecast. \

Finally, repeat the exercise above by combining the three kernels into one by multiplying them, instead of summing them up. Compare the results obtained in both cases and elaborate on why they may differ.

**Answer:** First we created a new column datetime to merge the date and time, then we eliminated the posterior values from the target datetime. Since temperature in a year is seasonal, we calculated only day differences and ignore the year by ignoring global warming since it has a minor effect. Moreover, we convert hour to 12 values. For example we assumed that 23.00 and 01.00 has 2 hour difference, not 22. \

The code is as follows:
```{r, warning=FALSE}
set.seed(1234567890)
stations <- read.csv("stations.csv", fileEncoding = "latin1")
temps <- read.csv("temps50k.csv")
st <- merge(stations,temps,by="station_number") # merged dataframe

# Create a new column with date and time
st$datetime <- as.POSIXct(paste(as.character(st$date), as.character(st$time)), 
                          format = "%Y-%m-%d %H:%M:%S", tz = "GMT")

# choose manually a width that gives large kernel values to closer points 
# and small values to distant points
h_distance <- 50
h_date <- 1
h_time <- 0.4

# Predict for each of these times
times <- c("04:00:00", "06:00:00", "08:00:00", "10:00:00", "12:00:00", 
           "14:00:00", "16:00:00", "18:00:00", "20:00:00", "22:00:00", 
           "24:00:00") 

gaussian_kernel <- function(diff_vector, l){ 
  kernel <- exp(-(diff_vector / (2*(l^2))))
  return(kernel)
}
```

```{r}
# Point to predict
a <- 	55.3836	#latitude
b <- 12.8203	 #longitude
date <- "1995-11-03"

i <- 1
target_datetime <- as.POSIXct(paste(date, times[i]), format = "%Y-%m-%d %H:%M:%S", tz = "GMT")
datetime_diff <- target_datetime - st$datetime
st_new <- st[which(datetime_diff > 0), ] # eliminate the posterior 

# distance calculation
distance_new <- distHaversine(cbind(st_new$latitude, st_new$longitude), c(a, b))

# date calculation
date1 <- as.Date(date, format = "%Y-%m-%d")
date2 <- as.Date(st_new$date, format = "%Y-%m-%d")

day_of_year1 <- as.numeric(format(date1, "%j"))
day_of_year2 <- as.numeric(format(date2, "%j"))

date_new <- abs(day_of_year2 - day_of_year1)

# hour calculation
st_new$hour <- as.numeric(format(as.POSIXct(st_new$datetime), "%I"))
target_hour <- as.numeric(substr(times[i], 1, 2))
time_new_init <- abs(st_new$hour - target_hour) 
time_new <- ifelse(time_new_init < 12, time_new_init, 24-time_new_init) # convert to 12 

# kernel calculation
distance_kernel <- gaussian_kernel(distance_new, h_distance)
date_kernel <- gaussian_kernel(date_new, h_date)
time_kernel <- gaussian_kernel(time_new, h_time)

kernel_diff <- c(gaussian_kernel(max(distance_new), h_distance),
                 gaussian_kernel(min(distance_new), h_distance),
                 gaussian_kernel(max(date_new), h_date),
                 gaussian_kernel(min(date_new), h_date),
                 gaussian_kernel(max(distance_new), h_time), 
                 gaussian_kernel(min(distance_new), h_time))
diff_name <- c("max distance", "min distance",
          "max date", "min date",
          "max time", "min time")

kernel_df <- data.frame("Difference" = kernel_diff)
rownames(kernel_df) <- diff_name
kable(kernel_df, caption = "Kernel Differences")
```

```{r, echo=FALSE, fig.align='center', fig.width=10}
par(mfrow = c(1, 2))
plot(distance_new, distance_kernel, main = "Distance-Kernel Graph",
     xlab = "Distance Difference", ylab = "Kernel")
grid()

plot(date_new, date_kernel, main = "Date-Kernel Graph",
     xlab = "Date Difference", ylab = "Kernel")
grid()
```

```{r, echo=FALSE, fig.align='center', fig.width=4, fig.height=3}
plot(time_new, time_kernel, main = "Time-Kernel Graph",
     xlab = "Time Difference", ylab = "Kernel")
grid()
```
The bandwith values are selected 50, 1 and 0.4 for distance, date and time respectively. From the graphs and table above it can be observed that kernel is low when distance is high and vice versa. Also kernel values vary between 0 and 1 which is important because otherwise predictions will be around average temperature value. \

The functions to predict temperature is as follows for summation and multiplication:
```{r}
summation_kernel_fnc <- function(target_date, a, b){
  temp <- numeric(length=length(times))
  for (i in 1:length(times)){
    target_datetime <- as.POSIXct(paste(target_date, times[i]), 
                                  format = "%Y-%m-%d %H:%M:%S", tz = "GMT")
    datetime_diff <- target_datetime - st$datetime
    st_new <- st[which(datetime_diff > 0), ] # eliminate the posterior 
    
    # distance calculation
    distance_new <- distHaversine(cbind(st_new$latitude, st_new$longitude), c(a, b))
    
    # date calculation
    date1 <- as.Date(target_date, format = "%Y-%m-%d")
    date2 <- as.Date(st_new$date, format = "%Y-%m-%d")
    
    day_of_year1 <- as.numeric(format(date1, "%j"))
    day_of_year2 <- as.numeric(format(date2, "%j"))
    
    date_new <- abs(day_of_year2 - day_of_year1)
    
    # hour calculation
    st_new$hour <- as.numeric(format(as.POSIXct(st_new$datetime), "%I"))
    target_hour <- as.numeric(substr(times[i], 1, 2))
    time_new_init <- abs(st_new$hour - target_hour) 
    time_new <- ifelse(time_new_init < 12, time_new_init, 24-time_new_init) # convert to 12
    
    weight_distance <- gaussian_kernel(distance_new, h_distance)
    weight_date <- gaussian_kernel(date_new, h_date)
    weight_time <- gaussian_kernel(time_new, h_time)
    
    w <- weight_distance + weight_date + weight_time
    total_weight <- sum(w)
    weighted_sum <- sum(w * st_new$air_temperature) 
    
    temp[i] <- weighted_sum/total_weight
  }
  return(temp)
}
```

```{r}
multiplication_kernel_fnc <- function(target_date, a, b){
  temp <- numeric(length=length(times))
  for (i in 1:length(times)){
    target_datetime <- as.POSIXct(paste(target_date, times[i]), 
                                  format = "%Y-%m-%d %H:%M:%S", tz = "GMT")
    datetime_diff <- target_datetime - st$datetime
    st_new <- st[which(datetime_diff > 0), ] # eliminate the posterior 
    
    # distance calculation
    distance_new <- distHaversine(cbind(st_new$latitude, st_new$longitude), c(a, b))
    
    # date calculation
    date1 <- as.Date(target_date, format = "%Y-%m-%d")
    date2 <- as.Date(st_new$date, format = "%Y-%m-%d")
    
    day_of_year1 <- as.numeric(format(date1, "%j"))
    day_of_year2 <- as.numeric(format(date2, "%j"))
    
    date_new <- abs(day_of_year2 - day_of_year1)
    
    # hour calculation
    st_new$hour <- as.numeric(format(as.POSIXct(st_new$datetime), "%I"))
    target_hour <- as.numeric(substr(times[i], 1, 2))
    time_new_init <- abs(st_new$hour - target_hour) 
    time_new <- ifelse(time_new_init < 12, time_new_init, 24-time_new_init) # convert to 12
    
    weight_distance <- gaussian_kernel(distance_new, h_distance)
    weight_date <- gaussian_kernel(date_new, h_date)
    weight_time <- gaussian_kernel(time_new, h_time)
    
    w <- weight_distance * weight_date * weight_time
    total_weight <- sum(w)
    weighted_sum <- sum(w * st_new$air_temperature) 
    
    temp[i] <- weighted_sum/total_weight
  }
  return(temp)
}
```

**Prediction 1:**
```{r, fig.align='center', echo=FALSE, fig.width=10}
predict_data <- st[which(st$air_temperature==min(st$air_temperature)),]
target_date <- predict_data$date
a <- predict_data$latitude
b <- predict_data$longitude
actual_temp <- predict_data$air_temperature
df_forecast <- data.frame("Time" = times, 
                          "sum_kernel_forecast" = summation_kernel_fnc(target_date, a, b),
                          "prod_kernel_forecast" = multiplication_kernel_fnc(target_date, a, b))

cat("Actual temperature is:", actual_temp, "at", predict_data$time, 
    "in location (", predict_data$latitude, ",", predict_data$longitude, ")\n")
kable(df_forecast, caption = "Forecast Low Temperature")
par(mfrow = c(1, 2))
plot(df_forecast$sum_kernel_forecast, type="o", xaxt="n", ylab = "Temperature", 
     xlab = " ", main = "Sum kernel")
axis(side = 1, at = 1:length(times), labels = times, las = 2) 
grid()

plot(df_forecast$prod_kernel_forecast, type="o", xaxt="n", ylab = "Temperature", 
     xlab = " ", main = "Production kernel")
axis(side = 1, at = 1:length(times), labels = times, las = 2) 
grid()
```
**Prediction 2:**
```{r, fig.align='center', echo=FALSE, fig.width=10}
predict_data <- st[11769,]
target_date <- predict_data$date
a <- predict_data$latitude
b <- predict_data$longitude
actual_temp <- predict_data$air_temperature
df_forecast <- data.frame("Time" = times, 
                          "sum_kernel_forecast" = summation_kernel_fnc(target_date, a, b),
                          "prod_kernel_forecast" = multiplication_kernel_fnc(target_date, a, b))

cat("Actual temperature is:", actual_temp, "at", predict_data$time, 
    "in location (", predict_data$latitude, ",", predict_data$longitude, ")\n")
kable(df_forecast, caption = "Forecast 1")
par(mfrow = c(1, 2))
plot(df_forecast$sum_kernel_forecast, type="o", xaxt="n", ylab = "Temperature", 
     xlab = " ", main = "Sum kernel")
axis(side = 1, at = 1:length(times), labels = times, las = 2) 
grid()

plot(df_forecast$prod_kernel_forecast, type="o", xaxt="n", ylab = "Temperature", 
     xlab = " ", main = "Production kernel")
axis(side = 1, at = 1:length(times), labels = times, las = 2) 
grid()
```
**Prediction 3:**
```{r, fig.align='center', echo=FALSE, fig.width=10}
predict_data <- st[26872,]
target_date <- predict_data$date
a <- predict_data$latitude
b <- predict_data$longitude
actual_temp <- predict_data$air_temperature
df_forecast <- data.frame("Time" = times, 
                          "sum_kernel_forecast" = summation_kernel_fnc(target_date, a, b),
                          "prod_kernel_forecast" = multiplication_kernel_fnc(target_date, a, b))

cat("Actual temperature is:", actual_temp, "at", predict_data$time, 
    "in location (", predict_data$latitude, ",", predict_data$longitude, ")\n")
kable(df_forecast, caption = "Forecast")

par(mfrow = c(1, 2))
plot(df_forecast$sum_kernel_forecast, type="o", xaxt="n", ylab = "Temperature", 
     xlab = " ", main = "Sum kernel")
axis(side = 1, at = 1:length(times), labels = times, las = 2) 
grid()

plot(df_forecast$prod_kernel_forecast, type="o", xaxt="n", ylab = "Temperature", 
     xlab = " ", main = "Production kernel")
axis(side = 1, at = 1:length(times), labels = times, las = 2) 
grid()
```
**Prediction 4:**
```{r, fig.align='center', echo=FALSE, fig.width=10}
predict_data <- st[which(st$air_temperature==max(st$air_temperature)),]
target_date <- predict_data$date
a <- predict_data$latitude
b <- predict_data$longitude
actual_temp <- predict_data$air_temperature
df_forecast <- data.frame("Time" = times, 
                          "sum_kernel_forecast" = summation_kernel_fnc(target_date, a, b),
                          "prod_kernel_forecast" = multiplication_kernel_fnc(target_date, a, b))

cat("Actual temperature is:", actual_temp, "at", predict_data$time, 
    "in location (", predict_data$latitude, ",", predict_data$longitude, ")\n")
kable(df_forecast, caption = "Forecast High Temperature")

par(mfrow = c(1, 2))
plot(df_forecast$sum_kernel_forecast, type="o", xaxt="n", ylab = "Temperature", 
     xlab = " ", main = "Sum kernel")
axis(side = 1, at = 1:length(times), labels = times, las = 2) 
grid()

plot(df_forecast$prod_kernel_forecast, type="o", xaxt="n", ylab = "Temperature", 
     xlab = " ", main = "Production kernel")
axis(side = 1, at = 1:length(times), labels = times, las = 2) 
grid()
```
**Prediction Winter:** Let's forecast the date "2000-02-06" in the coordinates (57.7213, 16.4683)
```{r, fig.align='center', echo=FALSE, fig.width=10}
target_date <- "2000-02-06"
a <- 57.7213
b <- 16.4683
actual_temp <- predict_data$air_temperature
df_forecast <- data.frame("Time" = times, 
                          "sum_kernel_forecast" = summation_kernel_fnc(target_date, a, b),
                          "prod_kernel_forecast" = multiplication_kernel_fnc(target_date, a, b))

kable(df_forecast, caption = "Forecast Winter")

par(mfrow = c(1, 2))
plot(df_forecast$sum_kernel_forecast, type="o", xaxt="n", ylab = "Temperature", 
     xlab = " ", main = "Sum kernel")
axis(side = 1, at = 1:length(times), labels = times, las = 2) 
grid()

plot(df_forecast$prod_kernel_forecast, type="o", xaxt="n", ylab = "Temperature", 
     xlab = " ", main = "Production kernel")
axis(side = 1, at = 1:length(times), labels = times, las = 2) 
grid()
```

**Prediction Summer:**  Now, let's forecast the date "2000-07-06" in the coordinates (57.7213, 16.4683)
```{r, fig.align='center', echo=FALSE, fig.width=10}
target_date <- "2000-07-06"
a <- 57.7213
b <- 16.4683
actual_temp <- predict_data$air_temperature
df_forecast <- data.frame("Time" = times, 
                          "sum_kernel_forecast" = summation_kernel_fnc(target_date, a, b),
                          "prod_kernel_forecast" = multiplication_kernel_fnc(target_date, a, b))

kable(df_forecast, caption = "Forecast Summer")

par(mfrow = c(1, 2))
plot(df_forecast$sum_kernel_forecast, type="o", xaxt="n", ylab = "Temperature", 
     xlab = " ", main = "Sum kernel")
axis(side = 1, at = 1:length(times), labels = times, las = 2) 
grid()

plot(df_forecast$prod_kernel_forecast, type="o", xaxt="n", ylab = "Temperature", 
     xlab = " ", main = "Production kernel")
axis(side = 1, at = 1:length(times), labels = times, las = 2) 
grid()
```
### Conclusion
We cannot say for sure which kernel -summation or multiplication- is better since we don't have a test data but we can make a conclusion using our intuition. It can be observed that temperature predictions is lower in nights and early mornings. Also temperature is lower in winter and higher in summer. By looking at the prediction 1 and 4, it can be concluded that multiplication kernel is better at predicting extreme values since its more sensitive. Multiplication operation emphasizes differences between extreme values more.


