---
title: "Computer lab 2 block 1"
author:
- Simge Cinar
- Duc Tran
- William Wiik
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
  pdf_document:
    fig_caption: yes
    number_sections: yes
geometry: top=100pt,bottom=100pt,left=68pt,right=66pt
subtitle: 732A99
header-includes:
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{longtable}
- \usepackage{caption}
- \usepackage{fancyhdr}
- \usepackage{titling}
- \usepackage[swedish, english]{babel}
- \renewcommand{\headrulewidth}{0pt}
- \renewcommand{\and}{\\}
- \pretitle{\centering\vspace{0cm}{\large Laboration report in Machine Learning \par}\vspace{4cm}\Huge\textbf}
- \posttitle{\vspace{1cm}\large\textbf{}\par}
- \preauthor{\centering\vspace{4cm}\normalsize}
- \postauthor{\par\vspace{3cm}}
- \predate{\centering{\normalsize Division of Statistics and Machine Learning \\ Department
  of Computer Science \\ Linköping University \par}}
- \postdate{\par\vspace{2cm}}
- \raggedbottom
---

<!-- <!-- Väljer språk till svenska för automatiska titlar -->
<!-- \selectlanguage{swedish} -->

<!-- Byter språket på figur- och tabellbeskrivningar till angivna namn -->
\captionsetup[table]{name = Table}


<!-- Anger sidnumreringens position -->
\fancyhf{}
\fancyfoot[C]{\thepage}
\pagestyle{fancy}

<!-- Tar bort sidnumrering för förteckningar och titelsidan -->
\pagenumbering{gobble}

<!-- Anger sidbrytning -->
\clearpage

<!-- Skapar en innehållsförteckning och anger djupet av rubrikerna som ska visas -->
\setcounter{tocdepth}{3}
\tableofcontents

<!-- Anger sidbrytning -->
\clearpage

<!-- Börjar sidnumreringen på sida 1 efter att alla förteckningar visats -->
\pagenumbering{arabic}
\setcounter{page}{1}

<!-- Börjar med kapitel 1 -->

```{r options, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(knitr)
library(dplyr)
library(neuralnet)
library(cowplot)
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 4.5,
  fig.height = 4,
  fig.align = "center")
```


# Assignment 3. Neural Networks


## 3.1 

Train a neural network to learn the trigonometric sine function. To do so, sample 500 points uniformly at random in the interval [0, 10]. Apply the sine function to each point. The resulting value pairs are the data points available to you. Use 25 of the 500 points for training and the rest for test. Use one hidden layer with 10 hidden units. You do not need to apply early stopping. Plot the training and test data, and the predictions of the learned NN on the test data. You should get good results. Comment your results.

```{r}

# The data
set.seed(1234567890)
data_runif <- runif(500, min = 0, max = 10)
data_df <- data.frame(runif = data_runif, sin = sin(data_runif))

tr_data <- data_df[1:25,]
te_data <-  data_df[26:500,]

# Random initialization of the weights in the interval [-1, 1], 20 weights + 11 bias = 31
winit <- runif(31, min = -1, max = 1)

 


```

```{r}

# Neural Network
nn_model_assignment_1 <- neuralnet(sin ~ runif, data = tr_data, hidden = c(10), startweights = winit)

predictions <- predict(nn_model_assignment_1, newdata = te_data) %>% as.vector()

```

```{r, fig.show="hold", fig.cap = "\\label{} Training data, test data and predictions"}

plot(tr_data, cex = 2, xlab = "runif[0,10]", ylab = "Sin(x)")
points(te_data, col = "blue", cex = 1)
points(te_data[,1], predictions, col = "red", cex = 1)
legend("bottomright", legend = c("Train", "Test", "Prediction"),
col = c("black", "blue", "red"), pch = 16)


```

The neural network have successfully learned the sin(x) pattern as the test plot and prediction plot follows each other very well.



\clearpage

## 3.2

In question (1), you used the default logistic (a.k.a. sigmoid) activation function, i.e. act.fct = "logistic". Repeat question (1) with the following custom activation functions: $h_1(x) = x, h_2(x) = max{0, x}$ and $h_3(x) = \text{ln}(1 + exp x)$ (a.k.a. linear, ReLU and softplus). See the help file of the neuralnet package to learn how to use custom activation functions. Plot and comment your results.


```{r}

# Three custom activation function
h1 <- function(x) x
h2 <- function(x) ifelse(x > 0, x, 0) # same as max(0,x)
h3 <- function(x) log(1 + exp(x))


# Neural Network
nn_model_h1 <- neuralnet(sin ~ runif, data = tr_data, hidden = c(10), act.fct = h1, startweights = winit)
nn_model_h2 <- neuralnet(sin ~ runif, data = tr_data, hidden = c(10), act.fct = h2, startweights = winit)
nn_model_h3 <- neuralnet(sin ~ runif, data = tr_data, hidden = c(10), act.fct = h3, startweights = winit)

predictions_h1 <- predict(nn_model_h1, newdata = te_data) %>% as.vector()
predictions_h2 <- predict(nn_model_h2, newdata = te_data) %>% as.vector()
predictions_h3 <- predict(nn_model_h3, newdata = te_data) %>% as.vector()



```


```{r, fig.show="hold", fig.cap = "\\label{} Training data, test data and predictions for first activation function (linear)"}

plot(tr_data, cex = 2, xlab = "runif[0,10]", ylab = "Sin(x)")
points(te_data, col = "blue", cex=1)
points(te_data[,1], predictions_h1, col = "red", cex = 1)
legend("bottomright", legend = c("Train", "Test", "Prediction"),
col = c("black", "blue", "red"), pch = 16)


```

In figure 2, we can see that the prediction is just a straight line and does not not fit the test data well. This is because the linear activation function can only represent linear mappings between input and output. Therefore, the prediction does not fit the test data well because the data is non-linear.

```{r, fig.show="hold", fig.cap = "\\label{} Training data, test data and predictions for second activation function (ReLu)"}

plot(tr_data, cex = 2, xlab = "runif[0,10]", ylab = "Sin(x)")
points(te_data, col = "blue", cex = 1)
points(te_data[,1], predictions_h2, col = "red", cex = 1)
legend("bottomright", legend = c("Train", "Test", "Prediction"),
col = c("black", "blue", "red"), pch = 16)



```

In figure 3, when we use ReLu as a activation function we get better result than the linear activation function, but still this is very bad fit to the test data.

```{r, fig.show="hold", fig.cap = "\\label{} Training data, test data and predictions for third activation function"}

plot(tr_data, cex = 2, xlab = "runif[0,10]", ylab = "Sin(x)")
points(te_data, col = "blue", cex=1)
points(te_data[,1], predictions_h3, col = "red", cex = 1)
legend("bottomright", legend = c("Train", "Test", "Prediction"),
col = c("black", "blue", "red"), pch = 16)



```

In figure 4, we use activation function softmax $\text{ln}(1 + exp x)$ and the prediction seems to follow the the test data well. This activation function has the best prediction between the three activation function we have used.




## 3.3

Sample 500 points uniformly at random in the interval [0, 50], and apply the sine function to each point. Use the NN learned in question (1) to predict the sine function value for these new 500 points. You should get mixed results. Plot and comment your results.


```{r}

# The data
set.seed(1234567890)
data_runif <- runif(500, min = 0, max = 50)
data_df <- data.frame(runif = data_runif, sin =  sin(data_runif))





predictions <- predict(nn_model_assignment_1, newdata = data_df) %>% as.vector()

```


```{r, fig.show="hold", fig.cap = "\\label{} Training data, test data and predictions with runif[0,50]"}


plot(tr_data, cex = 2, xlab = "runif[0,50]", ylab = "Sin(x)", ylim = c(-12, 2), xlim = c(0,50))
points(data_df, col = "blue", cex=1)
points(data_df[,1], predictions, col = "red", cex=1)
legend("bottomleft", legend = c("Train", "Test", "Prediction"),
col = c("black", "blue", "red"), pch = 16)

```

In figure 5, the predictions follows the test data well between values from 0 to 10.
This makes sense, because the NN learned in question (1) trained values between 0 to 10.

## 3.4

In question (3), the predictions seem to converge to some value. Explain why this happens. To answer this question, you may need to get access to the weights of the NN learned. You can do it by running *nn* or *nn*$*weights* where *nn* is the NN learned.


```{r}

nn_model_assignment_1$weights

sigmoid <- function(x){
return(1 / (1 + exp(-x)))
}

#Define weight matrix for layer 1
W1 <- nn_model_assignment_1$weights[[1]][[1]][2,]
# Define offset vector for layer 1
b1 <- nn_model_assignment_1$weights[[1]][[1]][1,]
# Get q: Test with input 100
q <- sigmoid(W1 * 200 + b1)
#
b2 <- nn_model_assignment_1$weights[[1]][[2]][1,]
W2 <- nn_model_assignment_1$weights[[1]][[2]][2:11,]
cat("The value will converge to:", W2%*%q + b2, "as the input increases")

```



## 3.5

Sample 500 points uniformly at random in the interval [0, 10], and apply the sine func-tion to each point. Use all these points as training points for learning a NN that tries to predict $x$ from $sin(x)$, i.e. unlike before when the goal was to predict $sin(x)$ from $x$. Use the learned NN to predict the training data. You should get bad results. Plot and comment your results. Help: Some people get a convergence error in this question. It can be solved by stopping the training before reaching convergence by setting threshold = 0.1.



```{r}


set.seed(1234567890)
data_runif <- runif(500, min = 0, max = 10)
data_df <- data.frame(runif = data_runif, sin = sin(data_runif))

tr_data <- data_df[1:25,]
te_data <-  data_df[26:500,]

# Random initialization of the weights in the interval [-1, 1], 20 weights + 11 bias
winit <- runif(31, min = -1, max = 1)


```



```{r}

# Neural Network
nn_model <- neuralnet(runif ~ sin, data = tr_data, hidden = c(10), startweights = winit,
                      threshold = 0.1)

predictions <- predict(nn_model, newdata = te_data) %>% as.vector()

```

```{r, fig.show="hold", fig.cap = "\\label{} Training data, test data and predictions, x ~ sin(x)"}

plot(tr_data, cex = 2, ylab = "runif[0,10]", xlab = "Sin(x)", ylim = c(-2, 11))
points(te_data, col = "blue", cex = 1)
points(predictions, te_data[,1], col = "red", cex = 1)
legend("topleft", legend = c("Train", "Test", "Prediction"),
col = c("black", "blue", "red"), pch = 16)

# 
# plot(tr_data[,2], tr_data[,1], cex = 2, ylab = "runif[0,10]", xlab = "Sin(x)")
# points(te_data[,2], te_data[,1], col = "blue", cex = 1)
# points(te_data[,1], predictions, col = "red", cex = 1)
# legend("bottomleft", legend = c("Train", "Test", "Prediction"),
# col = c("black", "blue", "red"), pch = 16)

```

In figure 6, we can see that the predicted values does not fit the a test data at all.
This is because the $sin(x)$ function can return same value for different $x$ values and therefore the NN gets confused.






